Discrete Quantum Theories and Computing

### 2018-08-04 dissertation.tex
dissertation.tex Exported from dissertation.lyx...

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{iuphd}
\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[all]{xy}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage[ backend=biber]{biblatex}
\addbibresource{prop.bib}
\ExecuteBibliographyOptions{sorting=none,maxbibnames=10,doi=false,isbn=false,url=false}
\DeclareUnicodeCharacter{0229}{\c{e}}
\DeclareUnicodeCharacter{00A0}{~}
\usepackage{mmap}
\usepackage{comment}
\usepackage{framed}
\usepackage[braket,qm]{qcircuit}
\usepackage{mathrsfs}
\usepackage{setspace}
\usepackage{dsfont}

@@

%\usepackage[basic]{complexity}
\newcommand{\lang}[1]{{\ensuremath{\mathsf{#1}}}}
\newcommand{\newlang}[2]{\newcommand{#1}{\lang{#2}}}
\newlang{\usat}{UNIQUE-SAT}

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fp}{\Fpx{p}}
\newcommand{\Fppx}[1]{\Fpx{{#1}^2}}
\newcommand{\Fpp}{\Fppx{p}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathComplex\mathbf{P}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{D}\mathComplex\mathbf{P}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}
\newcommand{\uf}{U_{\!f}}
\newcommand{\scalarPlus}{+}
\newcommand{\boolt}{\textsf{Bool}} 
\newcommand{\bfalse}{\texttt{\textbf{false}}}
\newcommand{\btrue}{\texttt{\textbf{true}}}
\newcommand{\dotprod}{dot product}
\newcommand{\Tr}{\ensuremath{\mathop{\mathrm{Tr}}\nolimits}}
\newcommand{\fnorm}[1]{\mathsf{N}\left({#1}\right)}

@@

\begin{document}
%data for Title Page

@@

\author{Yu-Tsung Tai}

@@

\maketitle
%data for Acceptance Page
\committeeMember{Amr A. Sabry , PhD} \committeeMember{Gerardo
Ortiz, PhD} \committeeMember{Dylan Paul Thurston, PhD} \committeeMember{Andrew
J. Hanson, PhD} \committeeMember{Shouhong Wang, PhD} \defensedate{Defense
Date} \acceptancepage

@@

\begin{dedication}I would like to dedicate my thesis to my parents,
Cheng-Tien Tai  and Feng-Ming Chang. Thanks for encouraging me to
study abroad and all your support during my Ph.D. study. \todo{Find
a way to include my parent's Chinese name. If I can find a way, also
include Amr's Arabic name.}\end{dedication}

\begin{acknowledgments}Thank Prof. Gerardo Ortiz, Prof. Amr Sabry,
and Prof. Andrew Hanson for publishing three papers \cite{geometry2013,DQT2014,THOS2017}
with me, and enormous other things. These papers become the main part
of my thesis, especially most of Chapter~\ref{chap:DQT=000026DQC}
is based on ``Geometry of discrete quantum computing''~\cite{geometry2013}
and ``Discrete quantum theories'' (DQT)~\cite{DQT2014}; most of
Chapter~\ref{chap:QIVPM} is based on ``Quantum Interval-Valued
Probability: Contextuality and the Born Rule'' (QIVPM) \cite{THOS2017}.
Together with Prof. Dylan Thurston and Prof. Shouhong Wang, thank
you for getting together to understand what I did and passing my dissertation
proposal Spring 2017. Also, thank Prof. Dylan Thurston for serving
in my advisory committee and my Tier 3 committee, with other advisory
committee members passing my Computer Science qualifying exam Spring
2016. Especially, my survey of real computation in the qualifying
exam inspires me the comparison of quantum computer and analog computer
in Chapter~\ref{chap:Introduction}. Chapter~\ref{chap:Introduction}
also contains the motivation in my application for Rethinking Foundations
of Physics 2017 workshop reviewed by Prof. Amr Sabry. Thank Prof.
Lawrence Moss for inviting me to present our results in the interdisciplinary
logic seminar and SICE theory seminar, co-chairing my Tier 3 committee,
writing an assessment letter for me, and drawing my attention on Prof.
Abramsky's paper~\cite{Abramsky2012}, which inspired me to merge
quantum probability with discrete quantum theories. Thank John Gardiner
for inspiring discussion on the difficulties merging them~\cite{Gardiner2014}.
These results are improved and motivated the transition from DQT to
QIVPM in Sec.~\ref{sec:Toward-IVPM}. Thank Prof. Tom Lewis for helping
me organize the unpublished deterministic quantum algorithms in the
term paper for SLST-T501 Academic Writing course as presented in Sec.~\ref{discretequantumtheoryI}.
Thank Traci Nagle for going through the literature review about finite
precision measurement and contextuality discussed for CSCI-Y790 Writing
and Editing course as presented in Sec.~\ref{sec:Quantum-Contextuality}.

Thank Elizabeth ``Betsy'' Merceron and Kexin Chen for helping me
pass TEPAIC exam.

Thank Hao-Chun Lee and Hsien-Ching Kao for discussing what I did in
the prospective of a general Ph.D. in Physics, and working later in
more computational industry.

Finally, thank Jin-Ru Yang for holding my hand and leading me here,
and hope I could be with you to wherever place until the death separating
us apart eventually.\end{acknowledgments}
\begin{abstract}
Our primary research interest is to build a quantum computing model
characterizing realistic quantum computers. While most of the quantum
computing models based on uncomputable numbers, that is, the continuum
of real numbers, most of the classical computers in our daily life
are digital instead of analog computers. This highlight the necessity
to investigate discrete models for quantum theory and computing. Specifically,
we start from replacing the continuum of complex numbers by the discrete
finite fields. Although we have fruitful results on their geometric
implications and computing powers, their probability models are still
not completely satisfactory. To address this issue, we further exploited
quantum interval-valued probability, and proved an imprecise version
of foundational results such as the Gleason and Kochen-Specker theorems.
\end{abstract}
\tableofcontents{}

@@

This marriage of quantum mechanics and computer science first envisioned
and popularized by Feynman has created an awkward, but opportune,
moment. The embarrassing dilemma was concisely described by Aaronson
with the following three statements \cite{AaronsonArkhipov2011,BFRDARW2013}:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:Textbook-quantum-mechanics}Textbook quantum mechanics
is correct. 
\item \label{enu:no-efficient-classical-factoring}There does not exist
an efficient classical factoring algorithm. 
\item \label{enu:extended-Church-Turing-Thesis}The extended Church-Turing
thesis — that probabilistic Turing machines can efficiently simulate
any physically realizable model of computation — is correct \cite{BernsteinVazirani1997,Kaye2007}. 
\end{enumerate}
There is overwhelming evidence to support each of these statements.
The theoretical framework of quantum mechanics \ref{enu:Textbook-quantum-mechanics}
has withstood decades of experimental confirmation. Entire industries
are founded on the assumption \ref{enu:no-efficient-classical-factoring}
that algorithms like RSA are secure and they also have withstood years
of attempted attacks \cite{boneh1999twenty,wiki:RSAFactoring}. Finally
the entire field of complexity theory in computer science which has
also withstood years of field testing rests, in essence, on assumption
\ref{enu:extended-Church-Turing-Thesis} \cite{Aaronson2005,Piccinini2015}.
And yet at least one of these three statements \emph{must be false}!
Indeed if there is a corresponding efficient classical factoring algorithm,
then we concede \ref{enu:no-efficient-classical-factoring}. If it
is correct Shor's efficient factoring algorithm is realizable, and
we can prove there is no efficient classical factoring algorithm,
we concede \ref{enu:extended-Church-Turing-Thesis}. Otherwise, if
we cannot implement Shor's algorithm no matter how hard we try, textbook
quantum mechanics, i.e., \ref{enu:Textbook-quantum-mechanics} may
need to be improved by a better theory. It is unlikely that there
will be a simple resolution to this awkward situation. It is more
likely that the resolution will emerge from deep and careful analyses
of the foundations of each field.

@@

The part of conventional quantum theory (CQT) used by quantum circuit
model is described by the following:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:-orthonormal-basis}$D$ orthonormal basis vectors for
a Hilbert space of dimension~$D$, 
\item \label{enu:probability-amplitude}the normalized $D$ complex probability
amplitude coefficients describing the contribution of each basis vector, 
\item \label{enu:unitary-matrix}a set of probability-conserving unitary
matrix operators that suffice to describe all required state transformations
of a quantum circuit, 
\item \label{enu:measurement-framework}and a measurement framework. 
\end{enumerate}
In Sec.~\ref{CQC1qubitBloch.sec}, we focus on the discrete geometric
issues raised by the properties \ref{enu:-orthonormal-basis} and
\ref{enu:probability-amplitude} given above for CQT. In Sec.~\ref{sec:Quantum-Circuit-Model},
we address the property~\ref{enu:unitary-matrix} by describing product
and entangled $n$-qubit states and unitary matrices in quantum circuit
model. In Sec.~\ref{sec:fuzzy}, we introduce the important issues
of \ref{enu:measurement-framework} and the foundations of quantum
probability space.

@@

\subsection{Two-dimensional Hilbert Space}

@@

We can reduce 3 degrees of freedom in \Eq{A5.eq} to 2 degrees of
freedom by effectively removing $\rme^{\rmi\theta}$ (``fibering
out by the circle $\Sphere{1}$''). The standard form of this maps
(``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{eq:HopfFibration}\\
Z & = & \left|\alpha_{0}\right|^{2}-\left|\alpha_{1}\right|^{2}={x_{0}}^{2}+{y_{0}}^{2}-{x_{1}}^{2}-{y_{1}}^{2}\,.\nonumber 
\end{eqnarray}
By denoting the three-dimensional vector~$\left(X,Y,Z\right)$ as
$\Hat{a}$, \Eq{A5.eq} implies these transformed coordinates obeying
\begin{equation}
\left\Vert \Hat{a}\right\Vert ^{2}=X^{2}+Y^{2}+Z^{2}=\left(\left|\alpha_{0}\right|^{2}+\left|\alpha_{1}\right|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point}~$\left(\phi,\psi\right)$ on the
surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

@@

The irreducible states in a $D$-dimensional Hilbert space are encoded
in a similar family of geometric structures known technically as the
complex projective space $\CP{D-1}$. We obtain these structures starting
with the $D$ initially unnormalized complex coefficients of the $D$-dimensional
basis $\ket{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}\ket{i}$. We then follow
the analog of the two-dimensional procedure: Conservation of probability
requires that the norm of the vector $\VVec{\alpha}$ be normalized
to unity: 
\begin{equation}
\braket{\Psi}{\Psi}=\|\VVec{\alpha}\|^{2}=\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=1\,.\label{probEq1.eq}
\end{equation}
Thus the initial equation for the geometry of a quantum state describes
a \textit{topological sphere~}$\Sphere{2D-1}$ embedded in $\mathReal^{2D}$.
To see this, remember that we can write the real and imaginary parts
of $\alpha_{i}$ as $\alpha_{i}=x_{i}+\rmi y_{i}$, so 
\begin{equation}
\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=\sum_{i=0}^{D-1}{x_{i}}^{2}+{y_{i}}^{2}=1
\end{equation}
describes the locus of a $2D$-dimensional real unit vector in $\mathReal^{2D}$,
which is by definition $\Sphere{2D-1}$, the $(2D-1)$-sphere.

@@

\subsection{Explicit generalization of the Hopf fibration construction}

@@

\section{Quantum Circuit Model\label{sec:Quantum-Circuit-Model}}

>>>>

%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[11pt,twoside]{iuphd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fontspec}
\setmainfont[Mapping=tex-text]{Times New Roman}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{booktabs}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{pdfpages}
\usepackage{graphicx}
\usepackage[all]{xy}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\providecommand{\LyX}{L\kern-.1667em\lower.25em\hbox{Y}\kern-.125emX\@}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{mmap}
\usepackage{comment}
\usepackage{framed}
\usepackage{mathrsfs}
\usepackage{setspace}
\usepackage{dsfont}
\usepackage{xeCJK}
\setCJKmainfont{新細明體}
\usepackage{unicode-math}

@@

%\usepackage[basic]{complexity}
\newcommand{\lang}[1]{{\ensuremath{\textsf{#1}}}}
\newcommand{\newlang}[2]{\newcommand{#1}{\lang{#2}}}
\newlang{\usat}{UNIQUE-SAT}
\newcommand{\ComplexityFont}[1]{%
{\ensuremath{\textsf{#1}}}%extra {} makes everyone happy.
}
\newcommand{\NP}{\ComplexityFont{NP}}

\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{prop}{Proposition}[chapter]
\newtheorem{thm}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem*{method}{Method}

\usepackage[braket,qm]{qcircuit}
\newcommand{\proj}[1]{\op{#1}{#1}}

\newcommand{\muB}{\ensuremath{\mu^{\mathrm{B}}}}
\newcommand{\muC}{\ensuremath{\mu^{\mathrm{C}}}}
\newcommand{\muF}{\ensuremath{\mu^{\mathrm{F}}}}
\newcommand{\barmuD}{\ensuremath{\bar{\mu}^{\mathrm{D}}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\eventsC}{\ensuremath{\events_{C}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fp}{\Fpx{p}}
\newcommand{\Fppx}[1]{\Fpx{{#1}^2}}
\newcommand{\Fpp}{\Fppx{p}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathComplex\mathbf{P}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{D}\mathComplex\mathbf{P}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}
\newcommand{\uf}{U_{\!f}}
\newcommand{\scalarPlus}{+}
\newcommand{\boolt}{\textsf{Bool}} 
\newcommand{\bfalse}{\texttt{\textbf{false}}}
\newcommand{\btrue}{\texttt{\textbf{true}}}
\newcommand{\dotprod}{dot product}
\newcommand{\Tr}{\ensuremath{\mathop{\mathrm{Tr}}\nolimits}}
\newcommand{\Psibar}{\overline{\Psi}}
\newcommand{\cardp}[2]{#1 \, \slash\!\! \slash \, #2}
\def\round{\mathop{\rm round}\nolimits}
\newcommand{\ps}{\texttt{+}}
\newcommand{\ms}{\texttt{-}}
\newcommand{\Hilb}{\mathcal{H}}
\newcommand{\coreBorn}{\ensuremath{\overline{\Hilb}}}
\def\C{{\mathbb{C}}}
\newcommand{\ultramodular}{\mathcal{M}}
\newcommand{\ultramodularL}[1][]{\ensuremath{\ultramodular^{L{#1}}}}
\newcommand{\ultramodularR}[1][]{\ensuremath{\ultramodular^{R{#1}}}}

\newcommand{\interval}[1]{{\normalfont\textsf{\textbf{#1}}}}
\newcommand{\imposs}{\interval{F}}
\newcommand{\necess}{\interval{T}}
\newcommand{\unknown}{\interval{U}}

\usepackage{xparse}
\NewDocumentCommand{\fnorm}{s m}{\mathsf{N}\IfBooleanT{#1}{^{-1}}\left({#2}\right)}

\usepackage{refcount}
% https://tex.stackexchange.com/questions/16866/how-can-i-use-footnotemark-with-a-ref-argument

@@

\usepackage{polyglossia}
\setdefaultlanguage[variant=american]{english}
\setotherlanguage{arabic}
\usepackage[style=numeric-comp,sorting=none,maxbibnames=10,doi=false,isbn=false,url=false]{biblatex}
\addbibresource{prop.bib}
\begin{document}
%data for Title Page

@@

\author{Yu-Tsung Tai\\
(戴淯琮)}

@@

\maketitle
%data for Acceptance Page
\committeeMember{Amr Sabry (\textarabic{عمرو صبري}), PhD} \committeeMember{Gerardo
Ortiz, PhD} \committeeMember{Dylan Paul Thurston, PhD} \committeeMember{Andrew
J. Hanson, PhD} \committeeMember{Shouhong Wang (汪守宏), PhD} \defensedate{Defense
Date} \acceptancepage

@@

\begin{dedication}I would like to dedicate my thesis to my parents,
Cheng-Tien Tai (戴振沺) and Feng-Ming Chang (張鳳鳴). Thanks for encouraging
me to study abroad and all your support during my Ph.D. study.\end{dedication}

\begin{acknowledgments}I want to thank Prof.~Gerardo Ortiz, Prof.~Amr
Sabry (\textarabic{عمرو صبري}), and Prof.~Andrew Hanson for publishing
three papers \cite{geometry2013,DQT2014,THOS2018} with me, and enormous
other things. These papers become the main part of my thesis, especially
most of Chapter~\ref{chap:DQT=000026DQC} is based on “Geometry
of discrete quantum computing”~\cite{geometry2013} and “Discrete
quantum theories” (DQT)~\cite{DQT2014}; most of Chapter~\ref{chap:QIVPM}
is based on “Quantum Interval-Valued Probability: Contextuality
and the Born Rule” (QIVPM) \cite{THOS2018}. Together with Prof.~Dylan
Thurston and Prof.~Shouhong Wang (汪守宏), I want to thank you for getting
together to understand what I did and passing my dissertation proposal
Spring 2017. I also thank Prof. Dylan Thurston for serving in my advisory
committee and my Tier 3 committee, with other advisory committee members
passing my Computer Science qualifying exam Spring 2016. Especially,
my survey of real computation in the qualifying exam inspired me the
comparison of quantum computer and analog computer in Chapter~\ref{chap:Introduction}.
Chapter~\ref{chap:Introduction} also contains the motivation in
my application for Rethinking Foundations of Physics 2017 workshop
reviewed by Prof. Amr Sabry (\textarabic{عمرو صبري}). I want to thank
Prof. Lawrence Moss for inviting me to present our results in the
interdisciplinary logic seminar and theory seminar in Computer Science,
co-chairing my Tier 3 committee, writing an assessment letter for
me, and drawing my attention on Prof.~Abramsky's paper~\cite{Abramsky2012},
which inspired me to merge quantum probability with discrete quantum
theories. And I thank John Gardiner for inspiring discussion on the
difficulties merging them~\cite{Gardiner2014}. These results are
improved and motivated the transition from DQT to QIVPM in Sec.~\ref{sec:Toward-IVPM}.
I also thank Prof.~Tom Lewis for helping me organize the unpublished
deterministic quantum algorithms in the term paper for SLST-T501 Academic
Writing course as presented in Sec.~\ref{discretequantumtheoryI}.
I want to thank Traci Nagle for going through the literature review
about finite precision measurement and contextuality discussed for
CSCI-Y790 Writing and Editing course as presented in Sec.~\ref{sec:Quantum-Contextuality}.
I also thank Elizabeth “Betsy” Merceron and Kexin “Casey”
Chen (谌可心) for teaching me English in general so that others can understand
me more easily.

\todo{Type a table for acknowledgment so that I would have to write
“I thank” so many times here?}Many people in both department
of Mathematics and Computer Science need to do something more to make
my double-major works smoothly. I need to thank all of them, especially
directors of graduate study in Mathematics, especially Prof. Christopher
Judge and Prof. Matthias Weber; directors of Ph.D. study in Computer
Science, especially Prof. Yuqing Melanie Wu (吴愈青) and Prof. Funda
Ergun.

Thank Kin Wai Chan for suggesting me the correct direction for Propsition~\ref{appH.sec},
Prof. 朱樺 for drawing my attention on Prof. Wan's book~\cite{Wan2006},
and Weihua Liu (刘伟华) for remaining me analysts call the right side
of the IVPM outer measure.

Thank Meng-Wei Chen (陳孟瑋) as my best friend and roommate in Bloomington.
Thank Hao-Chun Lee (李浩君) and Hsien-Ching Kao (高憲慶) for discussing
what I did in the prospective of a general Ph.D. in Physics, and working
later in more computational industry.

Finally, thank Jin-Ru Yang (楊謹如) for everything could not be written
here.\end{acknowledgments}

\begin{preface}
\begin{itemize}
\item The latest version of this thesis is in GitHub:
\[
\hspace{-4em}\textrm{https://github.com/yuttai/Contextuality-and-Nonlocality-in-Discrete-Quantum-Theory/blob/master/dissertation.pdf}
\]
\item Its source code is typeset using \LyX ~\cite{LyX} in:
\[
\hspace{-4em}\textrm{https://github.com/yuttai/Contextuality-and-Nonlocality-in-Discrete-Quantum-Theory/blob/master/dissertation.lyx}
\]
\item Any comments can be left in the Issue part of the GitHub Repository:
\[
\hspace{-4em}\textrm{https://github.com/yuttai/Contextuality-and-Nonlocality-in-Discrete-Quantum-Theory/issues}
\]
\end{itemize}
\end{preface}
\begin{abstract}
Our primary research interest is to build a quantum computing model
characterizing realistic quantum computers. While most of the quantum
computing models based on uncomputable numbers, that is, the continuum
of real numbers, most of the classical computers in our daily life
are digital instead of analog computers. This highlight the necessity
to investigate discrete models for quantum theory and computing. Specifically,
we start from replacing the continuum of complex numbers by the discrete
finite fields. Although we have fruitful results on their geometric
implications and computing powers, their probability models are still
not completely satisfactory. To address this issue, we further exploited
quantum interval-valued probability, and proved an imprecise version
of foundational results such as the Gleason and Kochen-Specker theorems.
\end{abstract}
\tableofcontents{}

@@

The marriage of quantum mechanics and computer science first envisioned
and popularized by Feynman has created an awkward, but opportune,
moment. This embarrassing dilemma was concisely summarized by Prof.~Aaronson
that one of the the following three statements is false \cite{AaronsonArkhipov2011,BFRDARW2013}:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:Textbook-quantum-mechanics}Textbook quantum mechanics
is correct. 
\item \label{enu:no-efficient-classical-factoring}There does not exist
an efficient classical factoring algorithm. 
\item \label{enu:extended-Church-Turing-Thesis}The extended Church-Turing
thesis — that probabilistic Turing machines can efficiently simulate
any physically realizable model of computation — is correct \cites{Bernstein:1993:QCT:167088.167097}{Kaye2007}. 
\end{enumerate}
There is overwhelming evidence to support each of these statements.
The theoretical framework of quantum mechanics \ref{enu:Textbook-quantum-mechanics}
has withstood decades of experimental confirmation. Entire industries
are founded on the assumption \ref{enu:no-efficient-classical-factoring}
that algorithms like RSA are secure and they also have withstood years
of attempted attacks \cite{boneh1999twenty,wiki:RSAFactoring}. Finally
the entire field of complexity theory in computer science which has
also withstood years of field testing rests, in essence, on assumption
\ref{enu:extended-Church-Turing-Thesis} \cite{Aaronson2005,Piccinini2015}.
And yet at least one of these three statements \emph{must be false}!
Indeed if there is a corresponding efficient classical factoring algorithm,
then we concede \ref{enu:no-efficient-classical-factoring}. If it
is correct that Shor's efficient factoring algorithm is realizable,
and we can prove there is no efficient classical factoring algorithm,
then we concede \ref{enu:extended-Church-Turing-Thesis}. Otherwise,
if we cannot implement Shor's algorithm no matter how hard we try,
textbook quantum mechanics, i.e., \ref{enu:Textbook-quantum-mechanics}
may need to be improved by a better theory. It is unlikely that there
will be a simple resolution to this awkward situation. It is more
likely that the resolution will emerge from deep and careful analyses
of the foundations of each field.

@@

The part of conventional quantum theory (CQT) used by quantum circuit
model is described by the following:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:-orthonormal-basis}$D$ orthonormal basis vectors for
a Hilbert space of dimension~$D$, 
\item \label{enu:probability-amplitude}the normalized $D$ complex probability
amplitude coefficients describing the contribution of each basis vector, 
\item \label{enu:unitary-matrix}a set of probability-conserving unitary
matrix operators that suffice to describe all required state transformations
of a quantum circuit, 
\item \label{enu:measurement-framework}and a measurement framework. 
\end{enumerate}
In Sec.~\ref{CQC1qubitBloch.sec}, we focus on the discrete geometric
issues raised by the properties \ref{enu:-orthonormal-basis} and
\ref{enu:probability-amplitude} given above for CQT. In Sec.~\ref{sec:fuzzy},
we introduce the important issues of \ref{enu:measurement-framework}
and the foundations of quantum probability space. In Sec.~\ref{sec:Quantum-Circuit-Model},
we address the property~\ref{enu:unitary-matrix} by describing product
and entangled $n$-qubit states and unitary matrices in quantum circuit
model.

@@

\subsection{Two-dimensional Hilbert Space\label{subsec:Two-dimensional-Hilbert-Space}}

@@

We can reduce 3 degrees of freedom in \Eq{A5.eq} to 2 degrees of
freedom by effectively removing $\rme^{\rmi\theta}$ (“fibering
out by the circle $\Sphere{1}$”). The standard form of this maps
(“the Hopf fibration”) is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{eq:HopfFibration}\\
Z & = & \left|\alpha_{0}\right|^{2}-\left|\alpha_{1}\right|^{2}={x_{0}}^{2}+{y_{0}}^{2}-{x_{1}}^{2}-{y_{1}}^{2}\,.\nonumber 
\end{eqnarray}
By denoting the three-dimensional vector~$\left(X,Y,Z\right)$ as
$\Hat{a}$, \Eq{A5.eq} implies these transformed coordinates obeying
\begin{equation}
\left\Vert \Hat{a}\right\Vert ^{2}=X^{2}+Y^{2}+Z^{2}=\left(\left|\alpha_{0}\right|^{2}+\left|\alpha_{1}\right|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point}~$\left(\phi,\psi\right)$ on the
surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

@@

The irreducible states in a $D$-dimensional Hilbert space are encoded
in a similar family of geometric structures known technically as the
complex projective space $\CP{D-1}$. We obtain these structures starting
with the $D$ initially unnormalized complex coefficients of the $D$-dimensional
basis $\ket{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}\ket{i}$. We then follow
the analog of the two-dimensional procedure: Conservation of probability
requires that the norm of the vector $\VVec{\alpha}$ be normalized
to unity: 
\begin{equation}
\braket{\Psi}{\Psi}=\left\Vert \VVec{\alpha}\right\Vert ^{2}=\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=1\,.\label{probEq1.eq}
\end{equation}
Thus the initial equation for the geometry of a quantum state describes
a \textit{topological sphere~}$\Sphere{2D-1}$ embedded in $\mathReal^{2D}$.
To see this, remember that we can write the real and imaginary parts
of $\alpha_{i}$ as $\alpha_{i}=x_{i}+\rmi y_{i}$, so 
\begin{equation}
\sum_{i=0}^{D-1}|\alpha_{i}|^{2}=\sum_{i=0}^{D-1}{x_{i}}^{2}+{y_{i}}^{2}=1
\end{equation}
describes the locus of a $2D$-dimensional real unit vector in $\mathReal^{2D}$,
which is by definition $\Sphere{2D-1}$, the $(2D-1)$-sphere.

@@

\subsection{Explicit generalization of the Hopf fibration construction\label{subsec:Explicit-generalization-of}}

@@
